Data-driven Reduction of Multiscale Stochastic Dynamical Systems \thanks{This work was supported by ....}

### 2014-09-01 reducible_SDE/paper/SIADS_draft.tex
Added more to ito calc stuff

To locally recover the slow variables, we need to build a metric that is invariant to the fast variables. 
%
Typically, this is done by averaging out the fast variables. 
%
We propose to do this using the Mahalanobis distance, which is a local PCA-type approach that inverts the local covariance matrix of the data, thereby creating a space with normalized diffusion terms.
%
It can be shown that the Mahalanobis distance locally approximates: 
\begin{equation} \label{eq:mahalanobis}
\| \mathbf{f}(Y_1) - \mathbf{f}(Y_2) \|^2_M = \| Y_1 - Y_2 \|^2_2 + \mathcal{O}(\| \mathbf{f}(Y_1) - \mathbf{f}(Y_2) \|^4_2)
\end{equation}
where the components of $Y$ have uncoupled noise with unit variance,
i.e., the variables $y_1, \dots, y_n$ are governed by the following SDEs:
\begin{equation} \label{eq:NIV_formulation}
dy_i = b(y) dt + dW_i
\end{equation}
and $Y = \begin{bmatrix} y_1 & y_2& \cdots & y_n \end{bmatrix}^T$.

In particular, for SDEs of the form in \eqref{eq:general_SDE}, consider the transformation
\begin{equation} \label{eq:general_rescale}
\begin{aligned}
x_i = y_i, \: &1 \le i \le m \\
x_i = \frac{y_i}{\sqrt{\epsilon}} \: &1 \le i \le m
\end{aligned}
\end{equation}

Then, \eqref{eq:general_SDE} becomes
\begin{equation} 
\begin{aligned}
dy_i &= b_i(y_1, \dots, y_n) dt + dW_i, \: 1 \le i \le m \\
dy_i &= -\frac{b_i(y_1, \dots, y_n)}{\sqrt{\epsilon}} dt + dW_i , \: m+1 \le i \le n
\end{aligned}
\end{equation}
where $b_i(y_1, \dots, y_n) = a_i (y_1, \dots, y_m, \frac{y_{m+1}}{\sqrt{\epsilon}}, \dots, \frac{y_n}{\sqrt{\epsilon}})$.
%
From \eqref{eq:general_rescale}, we can see that if $x_i = \mathcal{O}(1)$, then $y_i = \mathcal{O}({\sqrt{\epsilon}})$ for $ m+1 \le i \le n$, and 
\begin{equation}
\| Y_2 - Y_1 \|^2_2 = \sum_{i=1}^m \| y_{i,2} - y_{i,1} \|^2 = \sum_{i=1}^m \|x_{i,2} - x_{i,1} \|^2 + \epsilon \sum_{i=m+1}^n \|x_{i,2} - x_{i,1}\|^2
\end{equation}
%
Remarkably, this metric induces two important features. 
%
One, rescaling the data so that all variables have unit diffusion terms means that fast variables are collapsed and become ``epsilon" small, as observed in the second term in the right hand side. 
%
Two, the righthandside of \eqref{eq:mahalanobis} is invariant (to fourth order) to the function $\mathbf{f}$ and allows for the approximation of the Euclidean distance between samples of the slow variables in the ``decoupled" SDE space.

For the specific example \eqref{eq:fast_slow_SDE}, the change of variables gives the following system:
\begin{equation}
\begin{aligned}
dy_1 &= adt + dW_1\\
dy_2 &= -\frac{y}{\epsilon} dt +  dW_2
\end{aligned}
\end{equation}
Therefore, $y_2$ is a stochastic variable whose diffusion is unity, and 
\begin{equation}
\| X_1 - X_2 \|^2_2 = \|x_1 - x_2 \|^2_2 + \epsilon \|y_1 - y_2 \|^2_2.
\end{equation}

@@

\begin{equation}
\begin{aligned}
f( X_\tau) = & f( X_\rho) \\
&+ \sum_{j=1}^d \left. \frac{\partial f}{\partial x^j} \right|_{X_\rho}  \int_{\rho}^{\tau}  dW_s^j \\ &
+  \left. \left( \sum_{k=1}^d a^k \frac{\partial f}{\partial x^k} + \frac{1}{2} \sum_{j=1}^d \frac{\partial^2 f}{\partial {x^j}^2} \right) \right|_{X_\rho} \int_{\rho}^{\tau} ds + \sum_{j=1}^m \sum_{k=1}^m \int_{\rho}^{\tau} \int_{\rho}^{s_2} \frac{\partial^2 f}{\partial x^j \partial x^k} dW_{s_1}^j dW_{s_2}^k \\
& + \sum_{j=1}^d  \int_{\rho}^{\tau} \int_{\rho}^{s_2} \left( \sum_{k=1}^d a^k \frac{\partial^2 f}{\partial x^j \partial x^k} + \frac{1}{2} \sum_{k=1}^d \frac{\partial^3 f}{\partial x^j \partial {x^k}^2} \right) dW_{s_1}^j ds_2 \\
& + \sum_{j=1}^d  \int_{\rho}^{\tau} \int_{\rho}^{s_2} \left( \sum_{k=1}^d a^k \frac{\partial^ f}{\partial x^k \partial x^j} + \frac{1}{2} \sum_{k=1}^d  \frac{\partial^3 f}{\partial {x^k}^2 \partial x^j} \right) ds_1 dW_{s_2}^j  \\
& + \int_{\rho}^{\tau} \int_{\rho}^{s_2} \left( \sum_{j=1}^d \sum_{k=1}^d \left[ a^j \frac{\partial}{\partial x^j}\left( a^k \frac{\partial f}{\partial x^k} \right) + \frac{1}{2} \frac{\partial^2}{\partial {x^j}^2} \left( a^k \frac{\partial f}{\partial x^k} \right) + \frac{1}{2} a^j \frac{\partial^3 f}{\partial x^j \partial {x^k}^2} + \frac{1}{4} \frac{\partial^4 f}{\partial {x^j}^2 \partial {x^k}^2} \right] \right)  ds_1 ds_2
\end{aligned}
\end{equation}

To compute the Mahalanobis distance, we require $J J^T$ (estimate the covariance matrix from the data).
%
We do this using local simulation ``bursts''. 
%
The error in the covariance estimation is... (add equations).
%
The lowest-order error term depends on the curvature of $f$.

TODO: define error function for covariance estimation

Substituting our estimate of the covariance into \eqref{eq:taylor_expansion}, we obtain
\begin{equation}
\begin{aligned}
&\| X_2 - X_1 \|^2 \\
\approx &  \frac{1}{2} (Y_2 - Y_1)^T \left((J J^T)^{-1}(Y_1)  + (J J^T)^{-1}(Y_2) \right) (Y_2 - Y_1) \\
& - \frac{\delta t}{2} (Y_2 - Y_1)^T \left( A(Y_1) + A(Y_2) \right) (Y_2 - Y_1)\\
& + n \left( \left| K_1 K_2 \right| + \left| \frac{ K_2^2}{4} \right|  + \left| \frac{K_1 K_3}{3} \right|  \right) \| Y_2 - Y_1 \| ^4  \\
& + \mathcal{O} (\|Y_1 - Y_2 \|^6 ) 
\end{aligned}
\end{equation}
%
We want the first term in this expression to dominate.

TODO: interpret the different terms we get w.r.t. the curvature of f, the drift, the scale epsilon, and the window size.

From \eqref{eq:taylor_expansion}, we can see that the squared distance $\| X_2 - X_1 \|^2$, is a function of $\|Y_2 - Y_1\|$ and $\delta t$. 
%
If the first term in the expansion dominates, then $\|X_2 - X_1 \|^2$ should be independent of $\delta t$, and should vary quadratically with $\| Y_2 - Y_1\|$. 
%
We can find such a regime empirically (by plotting $\| X_2 - X_1 \|^2$ as a function of $\delta t$ and $\|Y_2 - Y_1\|$), and select a kernel distance where both conditions hold.

>>>>

To locally recover the slow variables, we need to build a metric that is invariant to the fast variables. 
%
Typically, this is done by averaging out the fast variables. 
%
We propose to do this using the Mahalanobis distance, which is a local PCA-type approach that inverts the local covariance matrix of the data, thereby creating a space with normalized diffusion terms.
%
It can be shown that the Mahalanobis distance locally approximates: 
\begin{equation} \label{eq:mahalanobis}
\| \mathbf{f}(\Xi_1) - \mathbf{f}(\Xi_2) \|^2_M = \| \Xi_1 - \Xi_2 \|^2_2 + \mathcal{O}(\| \mathbf{f}(\Xi_1) - \mathbf{f}(\Xi_2) \|^4_2)
\end{equation}
where the components of $\Xi$ have uncoupled noise with unit variance,
i.e., the variables $\xi_1, \dots, \xi_n$ are governed by the following SDEs:
\begin{equation} \label{eq:NIV_formulation}
d\xi_i = b(\xi) dt + dW_i
\end{equation}
and $\Xi = \begin{bmatrix} \xi_1 & \xi_2& \cdots & \xi_n \end{bmatrix}^T$.

In particular, for SDEs of the form in \eqref{eq:general_SDE}, consider the transformation
\begin{equation} \label{eq:general_rescale}
\begin{aligned}
x_i = \xi_i, \: &1 \le i \le m \\
x_i = \frac{\xi_i}{\sqrt{\epsilon}} \: &1 \le i \le m
\end{aligned}
\end{equation}

Then, \eqref{eq:general_SDE} becomes
\begin{equation} 
\begin{aligned}
d\xi_i &= b_i(\xi_1, \dots, \xi_n) dt + dW_i, \: 1 \le i \le m \\
d\xi_i &= -\frac{b_i(\xi_1, \dots, \xi_n)}{\sqrt{\epsilon}} dt + dW_i , \: m+1 \le i \le n
\end{aligned}
\end{equation}
where $b_i(\xi_1, \dots, \xi_n) = a_i (\xi_1, \dots, \xi_m, \frac{\xi_{m+1}}{\sqrt{\epsilon}}, \dots, \frac{\xi_n}{\sqrt{\epsilon}})$.
%
From \eqref{eq:general_rescale}, we can see that if $x_i = \mathcal{O}(1)$, then $\xi_i = \mathcal{O}({\sqrt{\epsilon}})$ for $ m+1 \le i \le n$, and 
\begin{equation}
\| \Xi_2 - \Xi_1 \|^2_2 = \sum_{i=1}^m \| \xi_{i,2} - \xi_{i,1} \|^2 = \sum_{i=1}^m \|x_{i,2} - x_{i,1} \|^2 + \epsilon \sum_{i=m+1}^n \|x_{i,2} - x_{i,1}\|^2
\end{equation}
%
Remarkably, this metric induces two important features. 
%
One, rescaling the data so that all variables have unit diffusion terms means that fast variables are collapsed and become ``epsilon" small, as observed in the second term in the right hand side. 
%
Two, the righthandside of \eqref{eq:mahalanobis} is invariant (to fourth order) to the function $\mathbf{f}$ and allows for the approximation of the Euclidean distance between samples of the slow variables in the ``decoupled" SDE space.

For the specific example \eqref{eq:fast_slow_SDE}, the change of variables gives the following system:
\begin{equation}
\begin{aligned}
d\xi_1 &= adt + dW_1\\
d\xi_2 &= -\frac{\xi}{\epsilon} dt +  dW_2
\end{aligned}
\end{equation}

@@

For our general multi scale + function SDE formulation,

For our general multi scale + function SDE formulation, 
\begin{equation}
\begin{aligned}
&\mathbb{E} \left[ f_1(X_\tau) f_2(X_\tau) \right] - \mathbb{E}[f_1(X_\tau)]\mathbb{E}[f_2(X_\tau)] = \\
& \sum_{i=1}^m \left. \frac{\partial f_1}{\partial x_i} \right|_{X_\rho} \left. \frac{\partial f_2}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left( \int_\rho^\tau dW_s^i \right)^2 \right] \\
& + \frac{1}{\epsilon} \sum_{i=m+1}^n \left. \frac{\partial f_1}{\partial x_i} \right|_{X_\rho} \left. \frac{\partial f_2}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left( \int_\rho^\tau dW_s^i \right)^2 \right] \\
& + \sum_{i=1}^m \left. \frac{\partial f_1}{\partial x_i} \right|_{X_\rho} \mathbb{E} \left[ \left( \int_\rho^\tau dW_s^i  \right) \left( \int_\rho^\tau \int_\rho^{s_2} \frac{\partial^2 f_2}{\partial x_i^2} dW_{s_1}^i dW_{s_2}^i  \right) \right] \\
& + \frac{1}{\epsilon^{3/2}} \sum_{i=m+1}^n \left. \frac{\partial f_1}{\partial x_i} \right|_{X_\rho} \mathbb{E} \left[ \left( \int_\rho^\tau dW_s^i  \right) \left( \int_\rho^\tau \int_\rho^{s_2} \frac{\partial^2 f_2}{\partial x_i^2} dW_{s_1}^i dW_{s_2}^i  \right) \right] \\
& + \sum_{i=1}^m \left. \frac{\partial f_2}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left( \int_\rho^\tau dW_s^i \right)  \left( \int_\rho^\tau \int_\rho^{s_2} \frac{\partial^2 f_1}{\partial x_i^2}  dW_{s_1}^i dW_{s_2}^i \right) \right] \\
& + \frac{1}{\epsilon^{3/2}} \sum_{i=m+1}^n \left. \frac{\partial f_2}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left( \int_\rho^\tau dW_s^i \right)  \left( \int_\rho^\tau \int_\rho^{s_2} \frac{\partial^2 f_1}{\partial x_i^2}  dW_{s_1}^i dW_{s_2}^i \right) \right] \\
&+ \mathcal{O} ((\tau - \rho)^2 ) 
\end{aligned}
\end{equation}

For the case when $f$ is linear,
\begin{equation}
\begin{aligned}
& \mathbb{E} [f_1(X_\tau) f_2(X_\tau) ] - \mathbb{E}[f_1(X_\tau)]\mathbb{E}[f_2(X_\tau)]= \\
=  &  \sum_{i=1}^m  \left. \frac{\partial f_1}{\partial x_i} \right|_{X_\rho} \left. \frac{\partial f_2}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left(\int_\rho^\tau dW_s^i \right)^2  \right] \\
& + \frac{1}{\epsilon} \sum_{i=m+1}^n  \left. \frac{\partial f_1}{\partial x_i} \right|_{X_\rho} \left. \frac{\partial f_2}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left(\int_\rho^\tau dW_s^i \right)^2  \right] \\
 & + \sum_{i=1}^m \left. \frac{\partial f_1}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left(\int_\rho^\tau dW_s^i \right)  \left(\int_\rho^\tau \int_{\rho}^{s_2} \left(  \sum_{k=1}^m \frac{\partial a_k}{\partial x_i} \frac{\partial f_2}{\partial x_k}  + \frac{1}{\epsilon} \sum_{k=m+1}^m \frac{\partial a_k }{\partial x_i} \frac{\partial f_2}{\partial x_k} \right)  dW_{s_1}^i ds_2 \right) \right] \\ 
 & + \frac{1}{\epsilon} \sum_{i=m+1}^n \left. \frac{\partial f_1}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left(\int_\rho^\tau dW_s^i \right)  \left(\int_\rho^\tau \int_{\rho}^{s_2} \left(  \sum_{k=1}^m \frac{\partial a_k}{\partial x_i} \frac{\partial f_2}{\partial x_k}  + \frac{1}{\epsilon} \sum_{k=m+1}^m \frac{\partial a_k }{\partial x_i} \frac{\partial f_2}{\partial x_k} \right)  dW_{s_1}^i ds_2 \right) \right] \\ 
 & + \sum_{i=1}^m  \left. \frac{\partial f_2}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left(\int_\rho^\tau dW_s^i \right)  \left(\int_\rho^\tau \int_{\rho}^{s_2} \left(  \sum_{k=1}^m \frac{\partial a_k}{\partial x_i} \frac{\partial f_1}{\partial x_k}  + \frac{1}{\epsilon} \sum_{k=m+1}^m \frac{\partial a_k }{\partial x_i} \frac{\partial f_1}{\partial x_k} \right) dW_{s_1}^i ds_2 \right) \right] \\
& + \frac{1}{\epsilon} \sum_{i=m+1}^n  \left. \frac{\partial f_2}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left(\int_\rho^\tau dW_s^i \right)  \left(\int_\rho^\tau \int_{\rho}^{s_2} \left(  \sum_{k=1}^m \frac{\partial a_k}{\partial x_i} \frac{\partial f_1}{\partial x_k}  + \frac{1}{\epsilon} \sum_{k=m+1}^m \frac{\partial a_k }{\partial x_i} \frac{\partial f_1}{\partial x_k} \right) dW_{s_1}^i ds_2 \right) \right] \\
& + \mathcal{O} ((\tau - \rho)^{5/2})
 \end{aligned}
\end{equation}

Therefore, we have
\begin{equation}
\begin{aligned}
e_c = &  \sum_{i=1}^m \left. \frac{\partial f_1}{\partial x_i} \right|_{X_\rho} \mathbb{E} \left[ \left( \int_\rho^\tau dW_s^i  \right) \left( \int_\rho^\tau \int_\rho^{s_2} \frac{\partial^2 f_2}{\partial x_i^2} dW_{s_1}^i dW_{s_2}^i  \right) \right] \\
& + \frac{1}{\epsilon^{3/2}} \sum_{i=m+1}^n \left. \frac{\partial f_1}{\partial x_i} \right|_{X_\rho} \mathbb{E} \left[ \left( \int_\rho^\tau dW_s^i  \right) \left( \int_\rho^\tau \int_\rho^{s_2} \frac{\partial^2 f_2}{\partial x_i^2} dW_{s_1}^i dW_{s_2}^i  \right) \right] \\
& + \sum_{i=1}^m \left. \frac{\partial f_2}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left( \int_\rho^\tau dW_s^i \right)  \left( \int_\rho^\tau \int_\rho^{s_2} \frac{\partial^2 f_1}{\partial x_i^2}  dW_{s_1}^i dW_{s_2}^i \right) \right] \\
& + \frac{1}{\epsilon^{3/2}} \sum_{i=m+1}^n \left. \frac{\partial f_2}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left( \int_\rho^\tau dW_s^i \right)  \left( \int_\rho^\tau \int_\rho^{s_2} \frac{\partial^2 f_1}{\partial x_i^2}  dW_{s_1}^i dW_{s_2}^i \right) \right] \\
&+ \mathcal{O} ((\tau - \rho)^2 ) 
\end{aligned}
\end{equation}
%
when $\frac{\partial f}{\partial x_i} \ne 0$, and 
%
\begin{equation}
\begin{aligned}
e_c =  &  \sum_{i=1}^m \left. \frac{\partial f_1}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left(\int_\rho^\tau dW_s^i \right)  \left(\int_\rho^\tau \int_{\rho}^{s_2} \left(  \sum_{k=1}^m \frac{\partial a_k}{\partial x_i} \frac{\partial f_2}{\partial x_k}  + \frac{1}{\epsilon} \sum_{k=m+1}^m \frac{\partial a_k }{\partial x_i} \frac{\partial f_2}{\partial x_k} \right)  dW_{s_1}^i ds_2 \right) \right] \\ 
 & + \frac{1}{\epsilon} \sum_{i=m+1}^n \left. \frac{\partial f_1}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left(\int_\rho^\tau dW_s^i \right)  \left(\int_\rho^\tau \int_{\rho}^{s_2} \left(  \sum_{k=1}^m \frac{\partial a_k}{\partial x_i} \frac{\partial f_2}{\partial x_k}  + \frac{1}{\epsilon} \sum_{k=m+1}^m \frac{\partial a_k }{\partial x_i} \frac{\partial f_2}{\partial x_k} \right)  dW_{s_1}^i ds_2 \right) \right] \\ 
 & + \sum_{i=1}^m  \left. \frac{\partial f_2}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left(\int_\rho^\tau dW_s^i \right)  \left(\int_\rho^\tau \int_{\rho}^{s_2} \left(  \sum_{k=1}^m \frac{\partial a_k}{\partial x_i} \frac{\partial f_1}{\partial x_k}  + \frac{1}{\epsilon} \sum_{k=m+1}^m \frac{\partial a_k }{\partial x_i} \frac{\partial f_1}{\partial x_k} \right) dW_{s_1}^i ds_2 \right) \right] \\
& + \frac{1}{\epsilon} \sum_{i=m+1}^n  \left. \frac{\partial f_2}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left(\int_\rho^\tau dW_s^i \right)  \left(\int_\rho^\tau \int_{\rho}^{s_2} \left(  \sum_{k=1}^m \frac{\partial a_k}{\partial x_i} \frac{\partial f_1}{\partial x_k}  + \frac{1}{\epsilon} \sum_{k=m+1}^m \frac{\partial a_k }{\partial x_i} \frac{\partial f_1}{\partial x_k} \right) dW_{s_1}^i ds_2 \right) \right] \\
& + \mathcal{O} ((\tau - \rho)^{5/2})
\end{aligned}
\end{equation}
